Evolutionary Algorithm for the Bonded Molecular Conformation Problem

### 2018-08-31 documentation/paper.tex
Tweaked SA parameters

Encoding each  molecule as a chain of two-dimensional cartesian coordinates
would require constraints to be enforced so that the relative orientations and
unit length bonds are maintained. In order to avoid this, therefore, molecules
will be encoded as sequences of $N-2$ angles, where $N$ is the number of atoms
in the molecule and each angle, $\alpha_i$ represents the rotation of the
segment following atom $a_i$. This encoding scheme means that cartesian
coordinates must be calculated at each Lennard-Jones potential computation as it
requires euclidean distances between pairs of atoms to be computed, but results
in much simpler crossover and mutation operations.

\subsection{Selection \& Crossover Method}

An elitist methodology will be utilised throughout the GA optimisation, meaning
that the single fittest configuration will be passed on unmodified to subsequent
generations in order to preserve beneficial traits and provide a mechanism for
remembering information about the best configuration that has been found so far.

Pairs of parents to be crossed over will then be selected using the Roulette
method, where the probability for a potential parent to be selected is based on
a function of its fitness value. The `elite' configurations that were inherited
from the previous generation will also be candidates for selection using this
method. Selected parents will undergo one of the following crossover operators
with equal probabilities of occurring:

\subsubsection{Single-Point Crossover}

A single atom $a_i$ will be randomly selected. Atoms to the left of this point
in one parent will form the left segment of the child, and the right segment of
the child will be similarly formed from the atoms to the right of the second
parent. A second child may be generated by repeating with the opposite segments
(the right segment from parent one and the left segment from parent two).
Finally, the angle $\alpha_i$ at the crossover point will be iterated through
discrete steps until an angle is found which results in the lowest energy, as
described by Pullan \cite{PULLAN1998331}.

\subsubsection{Two-Point Crossover}

Two crossover points will be selected at random and the molecules will be
combined similarly to the single-point crossover, with each child containing two
segments from one parent and one segment from the other. As with the single
point crossover, the angles of both crossover points will be optimised.

\subsection{Mutation and Intermediate Optimisation}

@@

Mutation will have a low chance of occurring so as not to corrupt fit
configurations too frequently and lose information provided by the global
search. Each generation, child nodes that are produced during the crossover
stage will have a small chance of becoming initial states in the simulated
annealing search.

\subsubsection{Annealing Schedule}

During annealing random angles are selected and modified by random amounts. If
the change produces an improved state, it is accepted. If the change produces a
worsened state, it is accepted with a probability of $1 - e^{\delta E / kT}$
such that the current `temperature' as well as how bad the state is determines
whether it is accepted.

The goal at this intermediate stage is not to find the global minimum, but
merely to provide a more localised search for deeper basins in the vicinity of
the current configuration; it is hoped that if the GA search has found a local
minimum in close proximity to the global minimum, the SA stage will take large
enough steps in order to climb over the boundary into the neighbouring basin
where it can be finally optimised by the gradient search in the next stage.

Therefore, an annealing schedule that prioritises speed over optimality will be
chosen.

@@

The final stage of the algorithm, after a candidate configuration has been found
by the GA and its immediate vicinity has been further explored through mutation,
is to refine the configuration using gradient-based local search. This stage
will use the gradient of the objective function in order to guarantee that the
bottom of the current basin will be reached. The quasi-Newton method
Broyden–Fletcher–Goldfarb–Shanno (BFGS) will be applied at this stage resulting
in a fast descent towards the current local minimum at each stage.

The coordinates of this locally optimised configuration will also be passed back
to the GA. This is known as a `Lamarkian' process and has been found to be
beneficial when used with this problem \cite{doi:10.1002/qua.560440214}.

>>>>

Encoding each  molecule as a chain of two-dimensional Cartesian coordinates
would require constraints to be enforced so that the relative orientations and
unit length bonds are maintained. In order to avoid this, therefore, molecules
will be encoded as sequences of $N-2$ angles, where $N$ is the number of atoms
in the molecule and each angle, $\alpha_i$ represents the rotation of the
segment following atom $a_i$. This encoding scheme means that Cartesian
coordinates must be calculated at each Lennard-Jones potential computation as it
requires euclidean distances between pairs of atoms to be computed, but results
in much simpler crossover and mutation operations.

\subsection{Selection \& Crossover Methods}

% An elitist methodology will be utilised throughout the GA optimisation,
% meaning that the single fittest configuration will be passed on unmodified to
% subsequent generations in order to preserve beneficial traits and provide a
% mechanism for remembering information about the best configuration that has
% been found so far.

% Pairs of parents to be crossed over will then be selected using the Roulette
% method, where the probability for a potential parent to be selected is based
% on a function of its fitness value. The `elite' configurations that were
% inherited from the previous generation will also be candidates for selection
% using this method. Selected parents will undergo one of the following
% crossover operators with equal probabilities of occurring:

The population is represented as a large pool of size $n_{pop} \times 2$ where
$n_{pop}$ is the user-provided population size. This pool contains both the
current population and the generated children in a single array such that it
can be easily sorted based on fitness.

% \subsubsection{Single-Point Crossover}

% A single atom $a_i$ will be randomly selected. Atoms to the left of this point
% in one parent will form the left segment of the child, and the right segment
% of the child will be similarly formed from the atoms to the right of the
% second parent. A second child may be generated by repeating with the opposite
% segments (the right segment from parent one and the left segment from parent
% two). Finally, the angle $\alpha_i$ at the crossover point will be iterated
% through discrete steps until an angle is found which results in the lowest
% energy, as described by Pullan \cite{PULLAN1998331}.

\subsubsection{Crossover}

A single-point crossover operator is applied as follows: A single angle index
$a_i$ is selected at random. Angles to the left of this point in the first
parent form the left segment of the child, and the right segment of the child is
similarly formed from the atoms to the right of the second parent.

\paragraph{Partial L-BFGS}

Finally, $\alpha_i$ is locally optimised by applying BFGS with this single angle
as its only optimisation parameter, with the rest of the angles fixed. Energy
and gradients for use in the BFGS algorithm are computed using the entire
molecule, while only the subset is optimised, leaving the rest of the angles
unchanged. This optimisation of the single crossover point effectively locally
optimises the relative orientation of the two connected segments.
%TODO: Should I leave this paragraph in?
Crossover operators with multiple crossover-points were not found to be
beneficial during experimentation and were even detrimental in some cases. As
such, they have been excluded from this study. The focus is primarily on
evaluating the use of annealing as a mutation operator and therefore the
crossover operation has been kept simple.

\subsubsection{Selection}

Pairs of parent configurations to be crossed over are selected by choosing
randomly from the fittest half of the population. This is repeated $n_{pop}$
times so that the number of children created is equal to the population size and
forms the next generation. Selecting from the fitter half of the population
ensures that information about the global search space is retained through
multiple generations by providing bias towards states that are closer to the
global minimum.

At each generation the entire pool is sorted so that the $n_{pop}$ fittest
states, including children and fitter parents from the current generation move
to the top half of the pool. This forms the new generation's population, and
implicitly provides a form of elitism by allowing states from the previous
generation to remain in the population if they are fit enough. After sorting,
the second half of the pool contains discarded states which can be later
overwritten with new children during subsequent crossover stages.

\subsection{Mutation}

@@

\subsubsection{Mutation Rate}

Mutation generally has a low chance of occurring so as not to corrupt fit
configurations too frequently and lose information provided by the global
search. It was found, however, that with a low mutation rate the search tended
to converge prematurely due to an overly homogeneous population. In order to
overcome this, the chance of mutation for each configuration is based on its
similarity to its direct neighbour. That is, each state is mutated if its
fitness is within one unit of that of the previous state.

Additionally, if a state's energy is positive it is guaranteed to be mutated.
This has the effect of reducing the number of knotted configurations in the
population, as the mutation operators tend to guide them towards a negative
energy state.

\subsubsection{Mutation Operators}

\paragraph{Unravelling}

This mutation operator is based on one described by Pullan \cite{PULLAN1998331}
in which A random angle index $\alpha_i$ is selected in the molecule and the
distance between this point and each endpoint is compared. The angles between
the selected angle and the closest end forms a new segment of angles. For each
angle in the selected segment, a very small (close to zero) random number is
chosen and the angle is set to that value. This results in one end of the
molecule being `unravelled' into a straight line, which effectively removes any
knots or kinks within that segment.The segment is later `rewound' through
subsequent local optimisation.

\paragraph{Guided Stochastic Mutation}

A random angle is selected and the current fitness and the current value of this
	angle is stored.The selected angle is set to a random value between $-\pi$
	and $\pi$ and the fitness is recomputed. If the new fitness is less than the
	previous fitness, revert the angle to its previous state. This process is
	then repeated until the maximum number of iterations is reached. This limit
	is important because after many generations it becomes increasingly
	difficult to find a random change in an angle that does not give lower
	energy, which would otherwise result in an infinite loop.

\paragraph{Naive Mutation}

\paragraph{Simulated Annealing Mutation}

A relatively brief annealing schedule is used due to the fact that during every
iteration an energy computation is required. The goal at this intermediate
search algorithm is not to find the global minimum, but merely to provide a more
localised search for deeper basins in the vicinity of the current configuration;
it is hoped that if the GA search has found a local minimum in close proximity
to the global minimum, the SA stage will take large enough steps in order to
climb over the boundary into the neighbouring basin where it can be finally
optimised by the gradient search in the next stage. Therefore, an annealing
schedule that prioritises speed over optimality will be chosen. The initial
temperature $T$ will be set to 100 and decrement by one every iteration until it
reaches 0.

During annealing, random angles are selected and modified by random amounts. If
the change produces an improved state, it is always accepted. If the change
produces a worsened state, it is accepted with a probability of $1 - e^{-\Delta
E / Tk}$ such that both the current `temperature' as well as the degree to which
the change will worsen the configuration, determine whether the change is to be
accepted. $\Delta E$ represents the change in energy that will occur if the new
configuration is accepted, $T$ represents the current temperature, and $k$ is
used to scale probability curve. A value of $k=0.03$ was chosen as this provides
a reasonable probability curve for energy deltas of 1 and below.

@@

% The final stage of the algorithm, after a candidate configuration has been
% found by the GA and its immediate vicinity has been further explored through
% mutation, is to refine the configuration using gradient-based local search.
% This stage will use the gradient of the objective function in order to
% guarantee that the bottom of the current basin will be reached. The
% quasi-Newton method Broyden–Fletcher–Goldfarb–Shanno (BFGS) will be applied at
% this stage resulting in a fast descent towards the current local minimum at
% each stage.

% The coordinates of this locally optimised configuration will also be passed
% back to the GA. This is known as a `Lamarkian' process and has been found to
% be beneficial when used with this problem \cite{doi:10.1002/qua.560440214}.

\subsection{GA Procedure and Parameters}

\section{Results}

\section{Discussion}

\section{Appendix}
